SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2019-01-24 scipy-1.0/paper.tex
Condense list-like roadmap mini-paragraphs into a Table.

% NOTE: some usage stats below are from https://libraries.io/pypi/scipy
% as well as github metrics; normally one doesn't put citations directly
% in abstract though (depends on field / journal)
\begin{abstract}
SciPy is an open source scientific computing library for the Python programming language.
SciPy 1.0 was released in late 2017, about 16 years after the original
version 0.1 release. SciPy
has become a \emph{de facto} standard for leveraging scientific algorithms
in the Python programming language, with more than 600
unique code contributors, millions of downloads per year, 161 dependent
packages, and 28700 dependent repositories.
The library includes functionality spanning clustering, Fourier transforms,
integration, interpolation, file I/O, linear algebra, image processing,
orthogonal distance regression, minimization algorithms, signal processing,
sparse matrix handling, computational geometry, and statistics. In this
work, we provide an overview of the capabilities and development practices of the
SciPy library and highlight some recent technical developments.
\end{abstract}
\begin{document}

@@

Yet, the philosophical motivations behind a fully open toolstack, and
an excited, friendly community with a singular focus, seemed to have
proven decisive in the long run.  It led not only to the library
described in this paper, but also to an entire ecosystem of related
packages, and a variety of social activities centered around them.  The
packages in the SciPy ecosystem share high standards of implementation,
documentation, and testing, and a culture eager to learn and adopt
better practices---both for community management and software
development.

@@

As STScI continued to use Python for an increasingly large portion
of the Hubble Space Telescope data analysis pipeline, they encountered
problems with the Python numerical array container.
Numeric, the original array package, was
suitable for small arrays, but not for the large images processed by
STScI.  With the Numeric maintainer's blessing, the decision was made
to write NumArray, a library that could handle data on a different
scale.  Unfortunately, NumArray proved inefficient for small arrays,
presenting the community with a rather unfortunate choice.  In 2005,
Travis Oliphant combined the best elements of Numeric and NumArray,
thereby solving the dilemma and paving the way for Python to become
a very significant player in the Data Science movement.\fixme{NumPy
  1.0 was released in October 2006}

@@

SciPy includes Cython\cite{behnel2011cython} wrappers for many BLAS and
LAPACK\cite{LAPACK} routines and for the special functions provided in the
\texttt{scipy.{\allowbreak}special} submodule.  These Cython wrappers are
available in the modules
\texttt{scipy.{\allowbreak}linalg.{\allowbreak}cython\_blas},
\texttt{scipy.{\allowbreak}linalg.{\allowbreak}cython\_lapack}, and
\texttt{scipy.{\allowbreak}special.{\allowbreak}cython\_special} respectively.
When writing algorithms in Cython, it is typically more efficient to call
directly into the libraries SciPy wraps rather than indirectly, using SciPy's
Python APIs.  These low-level interfaces for Cython can also be used outside of
the SciPy codebase to gain access to the functions in the wrapped libraries
while avoiding the overhead of Python function calls.  This can give
performance gains of one or two orders of magnitude for many use cases.

@@

\subsection*{Roadmap}

\texttt{fftpack} and {linalg} have too much functionality overlap with
\texttt{numpy} equivalents.

\texttt{scipy.interpolate} could benefit from a set of new features including
spline fitting routines with better user control; arithmetic routines for
splines; transparent tensor-product splines, Non-Uniform Rational B-Splines
support; and mesh refinement and coarsening of B-splines and corresponding
tensor products.  Some spline functionality in \texttt{scipy.signal} should be
migrated to \texttt{interpolate} and Second Order Sections need to be updated
to match the capabilities in other routines.

For \texttt{scipy.linalg}, the major change needed is to support a more recent
version of LAPACK. Users should be able access an increasing repertoire of
LAPACK functions using either \texttt{get\_lapack\_funcs} or
\texttt{sp.linalg.lapack.name\_of\_function}.

\texttt{scipy.optimize} would benefit from a few more good global optimizers 
as well as large-scale optimizers.

\texttt{scipy.ndimage} should be completely migrated to a data point
model (values on a grid), instead of a pixel model (elements with centers),
both of which have been used in the past, because the data point model is now
proving more effective in practice.

While \texttt{scipy.sparse} is mature, there are many data
formats and the community continues to move forward here so we will aim
to take advantage of their improvements\cite{abbasi2018sparse} 
while preserving backwards compatibility to the extent that it is possible.
\texttt{sparse.linalg} may benefit from wrappers for PROPACK 
for faster sparse SVD computation.

\texttt{scipy.spatial} will benefit from \texttt{spatial.transforms} with
support for rotation matrices.

\texttt{scipy.special} needs precision improvements for hypergeometric 
functions, parabolic cylinder functions, and spheroidal wave functions.

@@

From the user side, there were 13,096,468 downloads of SciPy from the Python
Packaging Index (PyPI) during the year 2017\cite{pypinfo}, which is a lower
bound on the number of total downloads by users given that downloading from
PyPI is only one of several popular methods to install SciPy.  The SciPy
website\cite{SciPylib}, which has been the default citation in the absence of a
peer-reviewed paper, has been cited $>3000$ times. Some of the most prominent
usages of or demonstrations of credibility for SciPy include the LIGO / Virgo
scientific collaboration that lead to the observation of gravitational waves
\cite{PhysRevLett.116.061102}, and the fact that SciPy is shipped directly with
MacOS and in the Intel Distribution for Python\cite{intel-python}.

\subsection*{Future development}

Future development plans include a major rewrite of the \texttt{scipy.sparse}
module, which is already underway\cite{abbasi2018sparse}. This is due to the
current interface only supporting two-dimensional objects. Even sub-arrays are
two-dimensional, and dimensions greater than two aren't supported. The new
package supports dimensions greater than two, and it bases itself around arrays
rather than matrices (which is strictly more powerful). It also features
improved inter-operability with NumPy and other community packages.

\fixme{This section should include key issues: sparse arrays, ndimage pixel vs
point, splines, fftpack vs. np.fft and linalg vs. np.linalg, under-maintained
submodules.}

@@

>>>>

% NOTE: some usage stats below are from https://libraries.io/pypi/scipy
% as well as github metrics; normally one doesn't put citations directly
% in abstract though (depends on field / journal)
\begin{abstract}
SciPy is an open source scientific computing library for the Python programming language.
SciPy 1.0 was released in late 2017, about 16 years after the original
version 0.1 release. SciPy
has become a \emph{de facto} standard for leveraging scientific algorithms
in the Python programming language, with more than 600
unique code contributors, millions of downloads per year, 161 dependent
packages, and 28700 dependent repositories. This includes usage of SciPy
in almost half of all machine learning projects on GitHub.
The library includes functionality spanning clustering, Fourier transforms,
integration, interpolation, file I/O, linear algebra, image processing,
orthogonal distance regression, minimization algorithms, signal processing,
sparse matrix handling, computational geometry, and statistics. In this
work, we provide an overview of the capabilities and development practices of the
SciPy library and highlight some recent technical developments.
\end{abstract}
\begin{document}

@@

Yet, the philosophical motivations behind a fully open toolstack, and
an excited, friendly community with a singular focus, seemed to have
proven decisive in the long run.  It led not only to the library
described in this paper, but also to an entire ecosystem of related
packages \cite{scipy-ecosystem}, and a variety of social activities centered around
them\cite{social-python}. The packages in the SciPy ecosystem 
share high standards of implementation,
documentation, and testing, and a culture eager to learn and adopt
better practices---both for community management and software
development.

@@

As STScI continued to use Python for an increasingly large portion
of the Hubble Space Telescope data analysis pipeline, they encountered
problems with the Python numerical array container.
Numeric, the original array package, was
suitable for small arrays, but not for the large images processed by
STScI.  With the Numeric maintainer's blessing, the decision was made
to write NumArray, a library that could handle data on a different
scale.  Unfortunately, NumArray proved inefficient for small arrays,
presenting the community with a rather unfortunate choice.  In 2005,
Travis Oliphant combined the best elements of Numeric and NumArray,
thereby solving the dilemma and paving the way for Python to become
a very significant player in the Data Science movement. NumPy
1.0 was released in October 2006\cite{numpy-1.0-tag}, with about 30
authors recognized for major contributions in its release notes.

@@

SciPy has provided special functions and leveraged BLAS and
LAPACK\cite{LAPACK} routines for many years. SciPy now additionally
includes Cython\cite{behnel2011cython} wrappers for:
many BLAS and LAPACK routines (added in 2015), and the special functions 
provided in the \texttt{scipy.{\allowbreak}special} submodule (added in 2016).  
These Cython wrappers are available in the modules
\texttt{scipy.{\allowbreak}linalg.{\allowbreak}cython\_blas},
\texttt{scipy.{\allowbreak}linalg.{\allowbreak}cython\_lapack}, and
\texttt{scipy.{\allowbreak}special.{\allowbreak}cython\_special} respectively.
When writing algorithms in Cython, it is typically more efficient to call
directly into the libraries SciPy wraps rather than indirectly, using SciPy's
Python APIs.  These low-level interfaces for Cython can also be used outside of
the SciPy codebase to gain access to the functions in the wrapped libraries
while avoiding the overhead of Python function calls.  This can give
performance gains of one or two orders of magnitude for many use cases.

@@

@@

From the user side, there were 13,096,468 downloads of SciPy from the Python
Packaging Index (PyPI) during the year 2017\cite{pypinfo}, which is a lower
bound on the number of total downloads by users given that downloading from
PyPI is only one of several popular methods to install SciPy.  The SciPy
website\cite{SciPylib}, which has been the default citation in the absence of a
peer-reviewed paper, has been cited $>3000$ times. Some of the most prominent
usages of or demonstrations of credibility for SciPy include the LIGO / Virgo
scientific collaboration that lead to the observation of gravitational waves
\cite{PhysRevLett.116.061102}, and the fact that SciPy is shipped directly with
MacOS and in the Intel Distribution for Python\cite{intel-python}. SciPy is used
by 47 \% of all machine learning projects on GitHub\cite{octoverse-scipy}.

\subsection*{Future Work}

Some additional roadmap items are summarized in Table~\ref{table:roadmap}.

\begin{table}[H]
\caption{Summary of SciPy Roadmap items following 1.0 release}
\begin{tabular}{ c|c }
    SciPy submodule & Summary of Change \\
    \hline
    \texttt{optimize} & a few more high quality global optimizers\\
    \texttt{fftpack} & reduce overlap with NumPy equivalent \\
    \texttt{linalg} & reduce overlap with NumPy equivalent \\
    \texttt{interpolate} & new spline fitting and arithmetic routines \\
    \texttt{interpolate} & new transparent tensor-product splines\\
    \texttt{interpolate} & new Non-Uniform Rational B-Splines\\
    \texttt{interpolate} & mesh refinement \& coarsening of B-splines and tensor products\\
    \texttt{signal} & migrate spline functionality to \texttt{interpolate}\\
    \texttt{signal} & Second Order Sections update to match capabilities in other routines\\
    \texttt{linalg} & support a more recent version of LAPACK\\
    \texttt{ndimage} & migrate to data point model from pixel model\\
    \texttt{sparse} & incorporate sparse arrays from Sparse package\cite{abbasi2018sparse} \\
    \texttt{sparse.linalg} & add PROPACK wrappers for faster SVD\\
    \texttt{spatial} & add support for (quaternion) rotation matrices\\
    \texttt{special} & precision improvements for hypergeometric, parabolic cylinder, and spheroidal wave functions\\
\end{tabular}\label{table:roadmap}
\end{table}

@@

\section*{Supporting Information}

% separate Figure numbering for SI
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{supporting_info/asv_bench/sparse/sparse_iteration_bench}
\caption{Airspeed velocity benchmarks for SciPy CSC and CSR matrix row iteration over roughly three years of commits prior to the release of version 1.0. Python version 2.7, Cython version 0.27.3, NumPy 1.8.2, and the indicated non-zero element density fractions were used. The partial performance regression observed shortly after the 1.0 release is likely related to a correctness fix\cite{sparse-regress}. The values were obtained from our regularly-updated suite of performance results available online at \url{http://pv.github.io/scipy-bench/}.}
\label{fig:sparse-iter}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{supporting_info/asv_bench/sparse/sparse_coo_dia_to_csr_csc}
\caption{Airspeed velocity benchmarks for SciPy COO / DIA to CSR / CSC matrix format conversion over roughly three years of commits prior to the release of version 1.0. Python version 2.7, Cython version 0.27.3, and NumPy 1.8.2 were used. The values were obtained from our regularly-updated suite of performance results available online at \url{http://pv.github.io/scipy-bench/}.}
\label{fig:sparse-conv}
\end{figure}
